### 2016-09-16 paper/5_evaluation.tex
Shorten evaluation section

%!TEX root=main.tex
\section{Evaluation}
\label{cha:evaluation}

In this chapter, we will investigate the scalability of Gilbert and its performance compared to famous hand-tuned ML algorithms.
We show that Gilbert is not only easily usable in terms of programming but also produces results with decent performance.
Furthermore, we compare the different execution engines and math-backends to see which system gives the best results.
Based on these outcomes, we want to come up with a recommendation for the best configuration of Gilbert.

@@

For our evaluation, we use the $400$-core cluster provided by the DIMA faculty of the Technical University of Berlin.
The cluster comprises $25$ local machines with 32 gigabytes of main memory per computer.
Each machine is equipped with $2$ AMD Opterons 6128 CPUs, each of them having $8$ cores.
The CPUs run at a speed of 2 gigahertz.

We employ Apache Spark-1.0.0~\cite{spark} for our test runs with the Spark execution engine.
For the Stratosphere execution engine, we use a slightly extended version of Stratosphere-0.6-SNAPSHOT~\cite{stratosphere}.
At the time of evaluation, Stratosphere-0.6 was still under development but we needed the latest features.
Therefore, we use the current snapshot version.
All extensions made to the current snapshot version are also pending pull requests and hopes are high that the final release will contain them all.
Thus, executing Gilbert with the stable release Stratosphere-0.6 should work perfectly fine.
As the underlying distributed file system both systems use Apache Hadoop-1.2.1~\cite{hadoop:2008a}.

The measured execution times on the cluster can slightly differ from measurement to measurement because of non-deterministic factors, such as cache misses, load caused by the operating system and network communication.
In order to cancel out the noise inflicted by these factors, we measure the execution time for each experimental setting five times and report the mean of the measurements.
That way, we reduce the overall variance of the measurements.

@@

The scalability evaluation investigates how Gilbert behaves under increasing work loads and how well it can exploit varying cluster sizes.
As we have implemented Gilbert to provide a scalable linear algebra environment, it is important that it can process data sizes exceeding the main memory of a single machine.
We want to evaluate the scalability performance for Stratosphere's and Spark's execution engine.

@@

As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ with $n$ being the dimensionality.
The matrix multiplication operation is demanding, both in CPU load as well as network I/O.
The implementation of the matrix multiplication, shown in \cref{sec:LinearAlgebraOperations}, first joins the column blocks of the left matrix with the row blocks of the right matrix.
This operation replicates the two operands partially and sends them across the network to their respective worker nodes.
For each matching pair of blocks, a local matrix multiplication is executed.
The local matrix multiplication has a complexity of $\mathcal{O}(n_{block}^3)$ with $n_{block}$ being the block size.
The produced intermediate results are grouped according to their left row and right column index.
Finally, the grouped result blocks are added up to give the final result.

The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells.
They are randomly generated prior to the matrix multiplication, thereby avoiding costly file system I/O.
The sparsity of both matrices is set to $0.001$.
Therefore, both matrices are represented by sparse matrices.
As a baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine.
The local execution engine uses the same math-backend as the distributed engines and thus serves as a good reference value to see the additional costs of parallel execution.
Breeze is chosen as the math back end for the matrix multiplication.
The Stratosphere and Spark execution engines are both started with 20 gigabytes of main memory for their task managers.
Furthermore, they are both configured to use a similar scheduling strategy, which distributes the work equally among the available computer nodes.
This aspect is especially important in order to compare the results, because some parts of Breeze are multi-threaded and, consequently, take advantage of idling cores.

\textbf{Increasing Problem Size}

In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $50$.
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior.
The resulting execution times for the local, Stratosphere and Spark execution engines are shown in \cref{fig:mmLoadRuntime}.

\begin{figure}[!h]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend entries={Local, Spark, Stratosphere},
				legend pos=north west,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadStratosphere};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Speedup Spark to Stratosphere},
				width=\dualpgfwidth,
				legend pos=north west,
				ymin=0.0,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Speedup,
			]
			{data/matrixMultLoad/matrixMultLoadSpeedup};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmSpeedup}
	\end{subfigure}
	\caption{Scalability of matrix multiplication on a $50$-core cluster. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmSpeedup} Speedup of Spark's execution engine compared to Stratosphere's.}
	\label{fig:mmBenchmark}
\end{figure}

On a single machine, we are able to execute the matrix multiplication for dimensionalities up to $n=10000$ before the system ran out of memory.
We can see that the local execution performs better for dimensionalities $n \le 5000$.
That is expected since the matrix still fits completely in the main memory of a single machine and the distributed execution adds some significant communication overhead.
Furthermore, Spark and Stratosphere both exhibit some noticeable job start up latency which is dominating the execution time for $n\le 10000$.
Those are the reasons why the local executor performs better than the distributed implementations for small matrix sizes.

For matrices with $n>5000$, Spark starts to calculate the matrix multiplication faster than the local executor.
The Stratosphere execution engine does not beat the local computation for sizes which are manageable by a single machine, though.
However, we can observe a steeper ascent of the local execution time compared to Stratosphere.

We also see that Gilbert can handle matrix sizes which scale far beyond the memory capacity of a single machine.
The matrix multiplication for $n=64000$ is finished in 719 seconds on Spark and in 1230 seconds on Stratosphere.
We can observe that the Spark execution engine runs consistently faster than Stratosphere.
Since the work load is similar for both systems, the difference has to be caused by the internal functioning of both systems.
For smaller dimensionalities, the execution time of both systems stays almost the same.
Therefore, it has to be mainly caused by the job start up times.
We can consequently infer that Spark is capable of starting its jobs faster than Stratosphere.

Not only does Spark starts its jobs faster, but it also performs better than Stratosphere without exception.
The speedup of Spark's execution engine compared to Stratosphere's execution engine calculating the matrix multiplication is given in \cref{fig:mmSpeedup}.
We can observe that there is a peak speedup of $7$ for $n=2500$.
However, we attribute this value to errors of measurement, because for $n>10000$ the speedup evens out at about $2$ with a slight decrease for $n=64000$.

\textbf{Increasing Cluster Size}

In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size.
As a benchmark, we calculate again $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ and $n$ being the dimensionality.
In order to observe the inflicted communication costs, we keep the work load per core constant while increasing the number of cores.
For this experiment we vary the number of cores from $1$ to $100$ and scale $n$ such that $n^3/\#cores$ is constant.
We started with $n=7500$ for a single core and reached $n=35000$ on $100$ cores.
As block size we chose $500\times 500$.
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.

\begin{figure}[!h]
	\centering
	\begin{tikzpicture}
		\begin{semilogxaxis}[
			xlabel={Number of cores},
			ylabel={Execution time $t$ in s},
			legend pos=north west,
			legend entries={Spark, Stratosphere},
			ymin=0.0,
			ymax=275,
			width=\dualpgfwidth,
		]
		\addplot[
			color=red,
			mark=o,
		] table[
			x=Parallelism,
			y=Time,
		]
		{data/matrixMultCores/matrixMultCoresSpark};
		\addplot[
			color=teal,
			mark=triangle,
		] table[
			x=Parallelism,
			y=Time,
		]
		{data/matrixMultCores/matrixMultCoresStratosphere};
		\end{semilogxaxis}
	\end{tikzpicture}
	\caption{Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmNodesRuntime}
\end{figure}

The optimal scale-out behavior would be a horizontal line.
However, it is impossible to achieve this scale-out behavior due to communication overhead inflicted by parallel execution.
Nonetheless, the results depicted in \cref{fig:mmNodesRuntime} indicate for both execution engines decent scale-out behavior.
Especially for the number of cores $\le 15$, we observe for Stratosphere and Spark an almost horizontal line.
From this point onwards, the scaling of Stratosphere degrades faster than Spark's scaling.
Spark exhibits an outstanding scale-out behavior.
The Spark execution engine requires 90 seconds to calculate the matrix multiplication with $n=35000$ on $100$ cores.
That is only a slowdown by a factor of $1.5$, if compared to the matrix multiplication with $n=7500$ on a single core, which takes 65 seconds to finish.
Interestingly, Spark's graph is not monotonic, as one would expect it.
For $\text{\#cores}=50$, we can observe a slight peak in the execution time.
This behavior is counterintuitive.
We suppose that this peculiarity is linked to the internal functioning of the Spark system.

@@

As second benchmark for evaluating the scalability properties, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V$ a factorization $W$ and $H$ such that $V\approx W H$ holds.
The algorithm is a popular ML algorithm which finds its application in computer vision, document clustering and topic modeling.
In the context of topic modeling, we would have $d$ documents and a set of $w$ words which are contained in the documents.
The goal of topic modeling is to identify the different topics and the words supporting a particular topic.
For this purpose, the matrix $V = (v_{i,j})_{i=1\ldots d,j=1\ldots w}$ is defined, with $v_{i,j}$ containing the frequency of a word $w_j$ appearing in document $d_i$.
By specifying the number of topics $t$, the GNMF algorithm computes $W\in \mathbb{R}^{d\times t}$ and $H\in \mathbb{R}^{t\times w}$ such that $V \approx W H$.
The row $w_i$ of $W$ indicates the topics the document $d_i$ contains and the row $h_i$ of $H$ says which words correlate with topic $t_i$.
The GNMF algorithm alternately updates the matrices $H$ and $W$ until the result converges.
The algorithm is given in \cref{lst:nmf}.
The operator $.*$, $./$ and $*$ denote the cell wise multiplication, the cell wise division and the matrix multiplication, respectively.

\begin{listing}[!h]
	\begin{CenteredBox}
		\begin{lstlisting}[language=Matlab,
			commentstyle=\color{black},
		  stringstyle=\color{black},
		  keywordstyle=\color{black}\bfseries,
		]
		V = load(); % load matrix to factorize
		W = load(); % load initial values of W
		H = load(); % load initial value of H

		while i < maxIterations
			H = H.*(W'*V ./ W'*W*H); % udpate H
			W = W.*(V*H' ./ W*H*H'); % update W
			i = i + 1;
		end
		\end{lstlisting}
	\end{CenteredBox}
	\caption{Non-negative matrix factorization algorithm.}
	\label{lst:nmf}
\end{listing}

We choose GNMF, because it is a ML algorithm which recently had been implemented for MapReduce systems~\cite{liu:2010a}.
As far as we know, the proposed MapReduce algorithm is one of the best distributed implementations of GNMF.
Thus, it is well suited to assess the performance of Gilbert's implementation of GNMF.

For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of documents $d$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
Thus, each line of $V$ contains roughly $100$ non-zero entries, which are drawn from a Gaussian distribution.
The matrices $W$ and $H$ are dense and initialized with random values drawn from a Gaussian distribution.
As a baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
As math back end we choose the Breeze library, which is also used for the distributed execution engines.
Like for the matrix multiplication benchmark, the task manager of Spark and Stratosphere are started with 20 gigabytes of memory and both systems use the same scheduling strategy keeping the work load on all nodes equally distributed.

\textbf{Increasing Problem Size}

In the first experiment we fix the number of cores to $50$ and investigate the runtime behavior for increasing values of $d$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to fairly compare the results with the optimized NMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using the Spark and Stratosphere runtime system.
This hand-tuned implementation, henceforth denoted as HT-GNMF, is also executed on $50$ cores.
Additionally, the data is generated having the respective partitioning required by the algorithm.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.

@@

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStepLoad/gnmfStepLoadSpeedupStratosphere};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:gnmfSpeedup}
	\end{subfigure}
	\caption{Scalability of GNMF on a $50$-core cluster. \subref{fig:nmfLoadRuntime} Execution time of one GNMF step depending on the data size. \subref{fig:gnmfSpeedup} Speedup of HT-GNMF running on Spark compared to Gilbert's GNMF using the Spark and Stratosphere execution engine, respectively.}
	\label{fig:nmfBenchmark}
\end{figure}

The local executor can be applied to sizes of $d$ ranging from $500$ to $1500$ rows, before the data exceeded the available main memory of a single computer.
As expected, the local execution performs far better for these data sizes than the distributed execution with Stratosphere.
However, the GNMF and the HT-GNMF executed on Spark outperformed the local execution.
That is rather surprisingly, since the distributed execution should inflict some noticeable communication overhead.
But apparently the parallelism compensates for the additional communication costs.

The distributed systems can also be used for data sizes which exceed the memory of a single computer.
Both distributed execution engines scale well up to the point where Spark and Stratosphere can no longer keep the data in memory.
The two data flow systems have to perform internal sorting, partitioning and shuffling steps to implement the high level operations, such as join, reduce, cogroup and cross.
If the memory size is not sufficient to execute these steps, then the data will be gracefully spilled to disk.
On the one hand, this behavior makes the systems more robust and applicable to data sizes which largely exceed the total amount of memory.
But on the other hand, the performance abruptly deteriorates massively once the data has to be spilled.
For the Stratosphere and Spark executor, we experienced this behavior for $d>50000$ and $d>150000$, respectively.
For data sizes bigger than these thresholds, the benchmark runs became so slow that we could not finish them and discarded these runs as infeasible.

For Stratosphere, this behavior ensues earlier due to its specific memory management.
Stratosphere assigns each cluster node a specific number of slots.
The default value is the number of cores.
The memory is then split evenly among the slots.
Once the task managers have started and assigned the memory to each slot, it is not possible to dynamically transfer memory portions between slots.
Therefore, the effectively available memory for each task is considerably smaller than the initial 20 gigabytes.
Additionally, Stratosphere's ability to support streaming further decreases the per task memory.
Currently, it is assumed that all tasks belonging to one pipeline can be deployed simultaneously to one slot.
In order to execute all pipeline tasks, the slot memory will be further divided by the number of tasks which can be concurrently run.

In contrast to that, Spark separates the execution of different operations into distinct stages.
A stage is only submitted for execution, after all preceding stages have been completed.
That way, the memory does not have to be splitted between succeeding tasks and thus spilling occurs later.

The runtime of HT-GNMF running on Spark, the Stratosphere and the Spark executor differ for varying data sizes almost by a constant factor.
The fastest implementation for $d\le 100000$ is the HT-GNMF algorithm running on Spark.
The speedup of HT-GNMF running on Spark compared to GNMF on Spark's and Stratosphere's executor is shown in \cref{fig:gnmfSpeedup}.
It can be seen that HT-GNMF runs approximately $1.7$ times faster than Gilbert's GNMF using the Spark executor.
Compared to the Stratosphere execution engine, we can observe that HT-GNMF runs faster and faster for increasing $d$.
HT-GNMF achieves a speedup of approximately $16$ for $d=50000$.
The different runtime behaviors of Spark's and Stratosphere's execution engines are most likely caused by the Spark and Stratosphere system, since the linear algebra operations are implemented identically.
The runtime behavior of HT-GNMF running on Stratosphere is interesting.
At first, it performs rather poorly compared to its Spark counterpart.
However, for $d>100000$ it suddenly outperforms the Spark implementation.
Thus, GNMF scales better if it is executed on Stratosphere.

Even though, Gilbert can not even reach the performance of HT-GNMF, the development using Gilbert was considerably easier.
One GNMF step can be programmed in five lines of Gilbert code, whereas we needed $28$ lines of Scala code for Spark's HT-GNMF and $70$ lines of Scala code for Stratosphere's HT-GNMF.
Not only did we have to know how to program Spark and Stratosphere, but it also took us quite some time to verify the correct functioning of both implementations.
The verification was made significantly more difficult and time-consuming due to a programming bug we introduced.
The debugging process showed us quite plainly how tedious the development process even with systems like Spark and Stratosphere can be.
Thus, the productivity increase gained by using a high-level declarative programming language for linear algebra must not be neglected and compensates for the performance loss.
\cite{alvaro:2010a} made a similar observations while developing a declarative programming language for distributed systems programming.

\textbf{Increasing Cluster Size}

In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $100$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $100000$ documents on $100$ cores.
The ideal behavior would be a horizontal line.
However, this outcome cannot be expected, since the GNMF computation requires communication between the cluster nodes.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.

\begin{figure}[h!]
	\centering
	\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				ymin=0,
				legend entries={Spark, Stratosphere},
				legend pos=north west,
				width=\dualpgfwidth,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStepCores/nnmfStepCoresSpark};

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStepCores/nnmfStepCoresStratosphere};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{Execution time of one GNMF step depending on the cluster size with constant work load per core.}
		\label{fig:nmfNodesRuntime}
\end{figure}

The scale-out behavior of the Stratosphere and Spark execution engines both show good results for $\#cores \le 5$.
For higher degrees of parallelism, Stratosphere's performance quickly deteriorates.
On $50$ cores, Stratosphere needs 1082 seconds, whereas Spark needs only 113 seconds.
We could not finish Stratosphere's computations for higher degrees of parallelism than $50$, because the system simply became too slow.
That vast performance decline is most likely caused by data spilling at some internal operation.
In contrast to Stratosphere, Spark does not suffer from these limitations for the number of cores we tested.
In fact, it almost exhibits an extraordinary scale-out behavior with an almost constant runtime.
The runtime on $100$ cores is only two times slower than the runtime on a single core with the same work load per core.

\subsection{Block Size}

The block size has a significant influence on the overall performance of Gilbert since it directly controls the data parallelism and data granularity of the local operations.
The bigger the block size is, the fewer blocks are available for parallel computations.
But the bigger the block size is, the more work can be done on a single computer without having to communicate with other nodes.
To measure the effect of the block size on the performance, we calculate a single GNMF step with varying block sizes using the Spark execution engine.
For this benchmark we set $d=100000$, $w = 100000$ and $t = 10$.
The remaining parameters are set like in \cref{subsec:NMF}.
The execution times for different block sizes are shown in \cref{fig:blocksizesNMFStep}.

>>>>

%!TEX root=paper.tex
\section{Evaluation}
\label{cha:evaluation}

In this chapter, we will investigate the scalability of Gilbert and its performance compared to famous hand-tuned ML algorithms.
We show that Gilbert is not only easily usable in terms of programming but also produces results with decent performance.

@@

For our evaluation, we use a $25$ machine cluster with 32 gigabytes of main memory per computer.
Each machine is equipped with $2$ AMD Opterons 6128 CPUs, each of them having $8$ cores.
The CPUs run at a speed of 2 gigahertz.

We employ Apache Spark-2.0.0~\cite{spark} for our test runs with the Spark execution engine.
For the Flink execution engine, we use Apache Flink-1.1.3~\cite{flink}.
As the underlying distributed file system both systems use Apache Hadoop-2.7.3~\cite{hadoop:2008a}.

All expirements have been executed five times and the mean of the measurement is reported.

@@

The scalability evaluation investigates how Gilbert behaves under increasing work loads and how well it can exploit varying cluster sizes.
As we have implemented Gilbert to provide a scalable linear algebra environment, it is important that it can process data sizes exceeding the main memory of a single machine.

@@

As a first benchmark, we choose the matrix multiplication $A\times B$ with $A,B \in \mathbb{R}^{n\times n}$ with $n$ being the dimensionality.
The matrix multiplication operation is demanding, both in CPU load as well as network I/O as it requires two network shuffles.

The matrices $A$ and $B$ are sparse matrices with uniformly distributed non-zero cells.
They are randomly generated prior to the matrix multiplication with a sparsity of $0.001$.
As a baseline, we run the same matrix multiplication on a single machine of the cluster using Gilbert's local execution engine.
The Flink and Spark execution engines are both started with 20 gigabytes of main memory for their task managers and they distribute the work equally among all worker nodes.

In the first experiment, we fixed the block sizes to $500 \times 500$ and set the number of cores to $50$.
We then increased the dimensionality $n$ of $A$ and $B$ to observe the runtime behavior.
The resulting execution times for the local, Flink and Spark execution engines are shown in \cref{fig:mmLoadRuntime}.

In the second experiment, we investigate the scaling behavior of Gilbert with respect to the cluster size.
In order to observe the inflicted communication costs, we keep the work load per core constant while increasing the number of cores.
For this experiment we vary the number of cores from $1$ to $100$ and scale $n$ such that $n^3/\#cores$ is constant.
We started with $n=7500$ for a single core and reached $n=35000$ on $100$ cores.
The results of the experiment are shown in \cref{fig:mmNodesRuntime}.

\begin{figure}[!h]
	\centering
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{loglogaxis}[
				xlabel={Dimensionality $n$},
				ylabel={Execution time $t$ in s},
				width=\dualpgfwidth,
				legend entries={Local, Spark, Flink},
				legend pos=north west,
			]
			\addplot[
				color=blue,
				mark=x,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadReference};
			\addplot[
				color=red,
				mark=o,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=RowsA,
				y=Time,
			]
			{data/matrixMultLoad/matrixMultLoadStratosphere};
			\end{loglogaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmLoadRuntime}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				legend pos=north west,
				legend entries={Spark, Stratosphere},
				ymin=0.0,
				ymax=275,
				width=\dualpgfwidth,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMultCores/matrixMultCoresSpark};
			\addplot[
				color=teal,
				mark=triangle,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/matrixMultCores/matrixMultCoresStratosphere};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:mmNodesRuntime}
	\end{subfigure}
	\caption{Scalability of matrix multiplication. \subref{fig:mmLoadRuntime} Execution time of matrix multiplication depending on the data size. \subref{fig:mmNodesRuntime} Execution time of matrix multiplication depending on the cluster size with constant work load per core.}
	\label{fig:mmScalability}
\end{figure}

On a single machine, we are able to execute the matrix multiplication for dimensionalities up to $n=10000$ before the system ran out of memory.
We can see that the local execution performs better for dimensionalities $n \le 5000$.
That is expected since the matrix still fits completely in the main memory of a single machine and the distributed execution adds some significant communication overhead and job start up latency.

For matrices with $n>5000$, Spark starts to calculate the matrix multiplication faster than the local executor.
The Flink execution engine does not beat the local computation for sizes which are manageable by a single machine, though.
We also see that Gilbert can handle matrix sizes which scale far beyond the memory capacity of a single machine.

The results depicted in \cref{fig:mmNodesRuntime} indicate for both execution engines decent scale-out behavior.
Especially for the number of cores $\le 15$, we observe for Flink and Spark an almost horizontal line.
From this point onwards, the scaling of Flink degrades faster than Spark's scaling.
The Spark execution engine requires 90 seconds to calculate the matrix multiplication with $n=35000$ on $100$ cores.
That is only a slowdown by a factor of $1.5$, if compared to the matrix multiplication with $n=7500$ on a single core, which takes 65 seconds to finish.

@@

As second benchmark for evaluating the scalability properties, we choose the Gaussian non-negative matrix factorization (GNMF) algorithm~\cite{seung:anips2001a}.
GNMF finds for a given matrix $V \in \mathbb{R}^{d\times w}$ a factorization $W \in \mathbb{R}^{d\times t}$ and $H \in \mathbb{R}^{t\times w}$ such that $V\approx W H$ holds.
For our evaluation, we calculate one step of the GNMF algorithm.
We set $t=10$, $w=100000$ and vary the number of rows $d$ of $V$.
The matrix $V\in\mathbb{R}^{d\times 100000}$ is a sparse matrix whose sparsity is set to $0.001$.
The non-zero cells of $V$ are uniformly distributed.
As a baseline, we run the GNMF on a single machine of the cluster using the local execution engine.
Like for the matrix multiplication benchmark, the task manager of Spark and Flink are started with 20 gigabytes of memory and both systems use the same scheduling strategy keeping the work load on all nodes equally distributed.

In the first experiment we fix the number of cores to $50$.
We start with $d=500$ and increase the number of rows of $V$ to $150000$.
The block size of Gilbert is set to $500 \times 500$.
In order to compare the results with the optimized NMF MapReduce implementation proposed in~\cite{liu:2010a}, we re-implemented the algorithm using the Spark and Flink runtime system.
This hand-tuned implementation, henceforth denoted as HT-GNMF, is also executed on $50$ cores.
The execution times of HT-GNMF and Gilbert's GNMF are shown in \cref{fig:nmfLoadRuntime}.

In the second experiment of the GNMF benchmark, we analyze how Gilbert scales-out when increasing the cluster size while keeping the work load for each core constant.
We vary the cluster size from $1$ core to $100$ cores and scale the number of documents $d$ accordingly.
Initially we start with $1000$ documents and, consequently, calculate the matrix factorization for $100000$ documents on $100$ cores.
The results of this experiment are shown in \cref{fig:nmfNodesRuntime}.

@@

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Rows,
				y=Speedup,
			]
			{data/nnmfStepLoad/gnmfStepLoadSpeedupStratosphere};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:gnmfSpeedup}
	\end{subfigure}
	\begin{subfigure}{\dualpgfwidth}
		\begin{tikzpicture}
			\begin{semilogxaxis}[
				xlabel={Number of cores},
				ylabel={Execution time $t$ in s},
				ymin=0,
				legend entries={Spark, Stratosphere},
				legend pos=north west,
				width=\dualpgfwidth,
			]
			\addplot[
				color=red,
				mark=o,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStepCores/nnmfStepCoresSpark};

			\addplot[
				color=teal,
				mark=diamond,
			] table[
				x=Parallelism,
				y=Time,
			]
			{data/nnmfStepCores/nnmfStepCoresStratosphere};
			\end{semilogxaxis}
		\end{tikzpicture}
		\caption{}
		\label{fig:nmfNodesRuntime}
	\end{subfigure}
	\caption{Scalability of GNMF on a $50$-core cluster. \subref{fig:nmfLoadRuntime} Execution time of one GNMF step depending on the data size. \subref{fig:gnmfSpeedup} Speedup of HT-GNMF running on Spark compared to Gilbert's GNMF using the Spark and Flink execution engine, respectively. \subref{fig:nmfNodesRuntime} Execution time of one GNMF step depending on the cluster size with constant work load per core.}
	\label{fig:nmfBenchmark}
\end{figure}

The local executor can be applied to sizes of $d$ ranging from $500$ to $1500$ rows, before the data exceeded the available main memory of a single computer.
As expected, the local execution performs far better for these data sizes than the distributed execution with Flink.
However, the GNMF and the HT-GNMF executed on Spark outperformed the local execution.
The distributed systems can also be used for data sizes which exceed the memory of a single computer.
Both distributed execution engines scale well up to the point where Spark and Flink can no longer keep the data in memory and have to spill it.

The runtime of HT-GNMF running on Spark, the Flink and the Spark executor differ for varying data sizes almost by a constant factor.
The fastest implementation for $d\le 100000$ is the HT-GNMF algorithm running on Spark.
The speedup of HT-GNMF running on Spark compared to GNMF on Spark's and Flink's executor is shown in \cref{fig:gnmfSpeedup}.
It can be seen that HT-GNMF runs approximately $1.7$ times faster than Gilbert's GNMF using the Spark executor.
Compared to the Flink execution engine, we can observe that HT-GNMF runs faster and faster for increasing $d$.
HT-GNMF achieves a speedup of approximately $16$ for $d=50000$.
The runtime behavior of HT-GNMF running on Flink is interesting.
At first, it performs rather poorly compared to its Spark counterpart.
However, for $d>100000$ it suddenly outperforms the Spark implementation.
Thus, GNMF scales better if it is executed on Flink.

Even though, Gilbert can not even reach the performance of HT-GNMF, the development using Gilbert was considerably easier.
One GNMF step can be programmed in five lines of Gilbert code, whereas we needed $28$ lines of Scala code for Spark's HT-GNMF and $70$ lines of Scala code for Flink's HT-GNMF.
Not only did we have to know how to program Spark and Flink, but it also took us quite some time to verify the correct functioning of both implementations.
The verification was made significantly more difficult and time-consuming due to a programming bug we introduced.
The debugging process showed us quite plainly how tedious the development process even with systems like Spark and Flink can be.
Thus, the productivity increase gained by using a high-level declarative programming language for linear algebra must not be neglected and compensates for the performance loss.
\cite{alvaro:2010a} made a similar observations while developing a declarative programming language for distributed systems programming.

The scale-out behavior of the Flink and Spark execution engines \cref{fig:nmfNodesRuntime} both show good results for $\#cores \le 5$.
For higher degrees of parallelism, Flink's performance quickly deteriorates.
On $50$ cores, Flink needs 1082 seconds, whereas Spark needs only 113 seconds.
That vast performance decline is most likely caused by data spilling at some internal operation.
The runtime on Spark with $100$ cores is only two times slower than the runtime on a single core with the same work load per core.

% \subsection{Optimizations}

% In this section, we want to evaluate the effects Gilbert's optimizer has on the runtime of programs.
% For this purpose, we execute one GNMF step on Gilbert's Stratosphere and Spark execution engine using the same settings as in \cref{subsec:NMF}.
% However, this time we disable the transpose pushdown and matrix-multiplication reordering optimization.
% The results are shown in \cref{fig:nnmfLoadOptimization}.

% \begin{figure}
% 	\centering
% 	\begin{subfigure}[t]{\dualpgfwidth}
% 		\begin{tikzpicture}
% 			\begin{semilogxaxis}[
% 				xlabel={Rows $d$ of $V$},
% 				ylabel={Execution time $t$ in s},
% 				legend pos=north west,
% 				legend entries={Optimized Spark, Optimized Stratosphere, Non-optimized Spark, Non-optimized Stratosphere},
% 				width=\dualpgfwidth,
% 			]

% 			\addplot[
% 				color=red,
% 				mark=o,
% 			] table[
% 				x=Rows,
% 				y=Time,
% 			]
% 			{data/nnmfStepLoadNoOp/nnmfStepLoadSpark};

% 			\addplot[
% 				color=teal,
% 				mark=triangle,
% 			]table[
% 				x=Rows,
% 				y=Time,
% 			]
% 			{data/nnmfStepLoadNoOp/nnmfStepLoadStratosphere};

% 			\addplot[
% 				color=blue,
% 				mark=x,
% 			] table[
% 				x=Rows,
% 				y=Time,
% 			]
% 			{data/nnmfStepLoadNoOp/nnmfStepNoOpLoadSpark};

% 			\addplot[
% 				color=black,
% 				mark=diamond,
% 			]table[
% 				x=Rows,
% 				y=Time,
% 			]
% 			{data/nnmfStepLoadNoOp/nnmfStepNoOpLoadStratosphere};

% 			\end{semilogxaxis}
% 		\end{tikzpicture}
% 		\caption{}
% 		\label{fig:nnmfLoadOptimization}	
% 	\end{subfigure}
% 	\begin{subfigure}[t]{\dualpgfwidth}
% 		\begin{tikzpicture}
% 			\begin{semilogxaxis}[
% 				xlabel={Rows $d$ of $V$},
% 				ylabel={Speedup},
% 				legend pos=north west,
% 				legend entries={Optimized Spark, Optimized Stratosphere},
% 				width=\dualpgfwidth,
% 			]

% 			\addplot[
% 				color=red,
% 				mark=o,
% 			] table[
% 				x=Rows,
% 				y=Speedup,
% 			]
% 			{data/nnmfStepLoadNoOp/gnmfStepLoadSpeedupSpark};
